Fluent statistical computing interfaces for biological data analysis

### 2020-03-23 Rmd/05-embeddings.Rmd
fleshing out para

A complimentary approach for visualizing structure in high dimensional data is
the tour. A tour is a sequence of projections of a high dimensional dataset
onto a low-dimensional orthonormal basis matrix, that is represented as a
dynamic visualization. The sequence of generated bases are interpolated to form
the tour path, allowing a user to explore the subspace of projections. A grand
tour corresponds to choosing new bases at random, and can give an overview of
the structure in the data. Instead of picking projections at random, a guided
tour can be used to generate a sequence 'interesting' projections as quantified
by an index function. Given the dynamic nature of the tour, user interaction is
important for controlling and exploring the visualisation: the tour has been
used previously by @Wickham2015-cx as tool for exploring statistical model fits
and by @Buja1996-fk for exploring factorial experimental designs.

@@

### Making comparisons: think global, act local

Aside from simple
linked brushing, we can also pose more complex queries via interrogating
the neighbourhood graph. Using a linked neighbourhood brush, we can 
visually investigate the nearest neighbour relationships in the 
high-dimensional space via brushing in the embedding view. The user can select
the number of nearest neighbours directly, and modify the distance metric 
used for determining the neighbours. This allows users to interrogate the stability of clusters generated in the embedding view. Multiple brushes
can be used to pose queries in either the tour view or the embedding view;
interface controls allow these brushes to combine using logical operators
such as 'and', 'or', or 'not'. The use of linked brushing goes beyond
simple color highlights, allowing analysts to get a more holistic
view of the effect of an embedding algorithm.

**distortion**: points that are far away in one view, will show neighbours that are close together in the opposing view.

**diffusion**: points that are close in one view, will show neighbours that are far apart in the opposing view.

>>>>

A complementary approach for visualizing structure in high dimensional data is
the tour. A tour is a sequence of projections of a high dimensional dataset
onto a low-dimensional orthonormal basis matrix, that is represented as a
dynamic visualization. The sequence of generated bases are interpolated to form
the tour path, allowing a user to explore the subspace of projections. A grand
tour corresponds to choosing new bases at random, and can give an overview of
the structure in the data. Instead of picking projections at random, a guided
tour can be used to generate a sequence 'interesting' projections as quantified
by an index function. Given the dynamic nature of the tour, user interaction is
important for controlling and exploring the visualisation: the tour has been
used previously by @Wickham2015-cx as tool for exploring statistical model fits
and by @Buja1996-fk for exploring factorial experimental designs.

@@

### Making comparisons: revising embeddings

As mentioned previously, when using any DR method, we are assuming
the embedding is representative of the high-dimensional dataset it was
computed from. Defining what it means for embedding to be 'representative` or
'faithful' to high-dimensional data is ill-posed and depends on the underlying
taks an analyst is trying to achieve. At the very minimum, we are interested
in distortions and diffusions of the high-dimensional data. Distortions
occur when points that are near each other in the embedding view are 
far from each other in the original dataset. This implies that the embedding 
is not continuous. Diffusions occur when points are far from each other in the embedding view are near in the original data.  Whether, points are near or far 
is reliant on the distance metric used; distortions and diffusions can
be thought of as the preservation of distances or the nearest neighbors
graphs between the high-dimensional space and the embedding space. As distances
can be noisy in high-dimensions, ranks can be used instead as has been 
proposed by ... Identifying distortions and diffusions allows an analyst
to investigate the quality of their embedding and revise them iteratively. We
propose this can be done visually using our side-by-side tour and embedding
views. Some quality checks, such as finding distortions can be identified using simple linked brushing, however we can also interrogate them via spatial
brushes and brush compostions.

Using a linked neighbourhood brush, we can 
visually investigate the nearest neighbour relationships in the 
high-dimensional space via brushing in the embedding view. The user can select
the number of nearest neighbours directly, and modify the distance metric 
used for determining the neighbours. This allows users to interrogate the stability of clusters generated in the embedding view. Multiple brushes
can be used to pose queries in either the tour view or the embedding view;
interface controls allow these brushes to combine using logical operators
such as 'and', 'or', or 'not'. The use of linked brushing goes beyond
simple color highlights, allowing analysts to get a more holistic
view of the effect of an embedding algorithm.
